---
title: "Lab 4"
author: Haochi Kiang (haoki222)
geometry: margin=0.8cm
output: pdf_document
---

```{r echo = F}
knitr::opts_chunk$set(fig.width=6.5, fig.height=8)

library(grid)
library(gridExtra)
library('lattice')
library('latticeExtra')
library('functional')
library('matrixStats') ## For numerically stable logSumExp

sumnorm = function (x) { print(x); x / sum(x) }

## Returns the log emission probability
demit = Vectorize(function(x, z, emit_sd) {
    lndens = dnorm(x, z + (-1:1), emit_sd, log = T)
    logSumExp(lndens) - log(3)
}, vectorize.args = 'z')

rtrans = Vectorize(function(z) {
    rnorm(1, z + (0:2)[sample(3,1)], 1)
})

sim_myssm = function(duration, emit_sd) {
    res = list()
    res$z = Reduce( function (lastz, mv) {
        c(lastz, rnorm(1, tail(lastz,1) + (0:2)[mv], 1))
    }, replicate(duration-1, sample(3,1)), init=runif(1,0,100))

    res$x = sapply( res$z, function (z) {
        rnorm(1, z + (-1:1)[sample(3,1)], emit_sd)
    })
    res
}

ahead_measure = function (cur_x, prev_measr, demitfn) {
    ## We don't need to normalize it here because sample() behaves the same
    ln_wgt = demitfn(cur_x, prev_measr)
    wgt = exp(ln_wgt - max(ln_wgt) + 10)
    rtrans(prev_measr[replicate(length(prev_measr),
                                sample(length(wgt), 1, replace=T, prob=wgt))])
}

ahead_measure_nowgt = function (cur_x, prev_measr, demitfn) {
    ## We don't need to normalize it here because sample() behaves the same
    wgt = rep(1, length(prev_measr))
    rtrans(prev_measr[replicate(length(prev_measr),
                                sample(length(wgt), 1, replace=T, prob=wgt))])
}


plot_history = function (measr_history, x, z) {
    timemax = ncol(measr_history)
    nsamp = nrow(measr_history)
    xy = do.call(cbind, lapply(1:timemax, function (tm)
        sapply(1:nsamp, function (s) c(tm, measr_history[s,tm]))))
    allplot = xyplot(x ~ seq_along(x), type='l', col=4, ylab='',
                     xlab = 'Time', main = 'All steps',
                     key = list(corner = c(0,0.98),
                                lines = list(col = c('blue','red','green', rgb(0,0,0,0.5)),
                                             type = c('l','l','l','p'), pch = 20),
                                text = list(c('Observation', 'Truth', 'Posterior Mean',
                                              'Particles')))) +
        as.layer(xyplot( xy[2,] ~ xy[1,], col=rgb(0,0,0,0.02),
                        pch = 20, cex = 0.5)) +
        as.layer(xyplot(z ~ seq_along(z), type='l', col=2)) +
        as.layer(xyplot( colMeans(measr_history) ~ 1:ncol(measr_history),
                        type='l', col='green'))
    beginplot = histogram(measr_history[,1], nint = 15, col = 1, ylab='',
                          main = 'Particles before any observation',
                          xlab = 'Particle location')
    secondplot = histogram(measr_history[,2], nint = 15, col = 1, ylab='',
                           main = 'Particles after one observation',
                           xlab = 'Particle location')
    thirdplot = histogram(measr_history[,3], nint = 15, col = 1, ylab='',
                          main = 'Particles after two observations',
                          xlab = 'Particle location')
    endplot = histogram(measr_history[,ncol(measr_history)],
                        nint = 15, col=1, ylab = '',
                        main='Final one-step-ahead prediction',
                        xlab = 'Particle location')

    grid.arrange(
        grobs = list(beginplot, secondplot,
                     thirdplot, endplot,
                     allplot),
        layout_matrix = rbind(c(1,2),
                              c(3,4),
                              c(5,5),
                              c(5,5)))
}

sim_and_plot = function (emit_sd, nsamp=100, ahead_measr = ahead_measure) {
    demit1 = Curry(demit, emit_sd = emit_sd)
    modsamp1 = sim_myssm(100, emit_sd)
    measr_history = matrix(NA, nsamp, length(modsamp1$x)+1)
    measr_history[,1] = runif(nsamp, 0, 100)
    for (i in seq_len(length(modsamp1$x)) ) {
        measr_history[,i+1] =
            ahead_measr(modsamp1$x[i], measr_history[,i], demit1)
    }
    plot_history(measr_history, modsamp1$x, modsamp1$z)

}

lattice.options(
    layout.heights=list(bottom.padding=list(x=0),
                        top.padding=list(x=0.4),
                        main.key.padding = list(x=-1),
                        axis.xlab.padding = list(x=0)),
  layout.widths=list(left.padding=list(x=-0.5), right.padding=list(x=-1))
)

```

## Implementation

For implementation of the algorithm please see Appendix.


## Emission standard deviation = 1

The following plot shows the particles at the first, second, third and
the last step, as well as an overview of the entire process in a line
plot. Emission standard deviation is set to 1 in this section.

The histogram of the particles at the beginning shows the state of the
particles before observing any data points. In other words, according to
the given model specification, this is uniform distribution with range
between zero and one hundred.

We can dr

```{r echo = F}
set.seed(1111)
sim_and_plot(1)

```

## Emission standard deviation = 5

```{r echo = F}
set.seed(1111)
sim_and_plot(5)

```

## Emission standard deviation = 50

```{r, echo = F}
set.seed(1111)
sim_and_plot(50)

```

## Always-constant weight, emission 

```{r}
set.seed(1111)
sim_and_plot(50, ahead_measr = ahead_measure_nowgt)

```



\newpage
## All codes
```{r ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
